---
title: "HW2_ozone"
author: "Darrell Sonntag"
date: "2024-01-20"
output: html
knitr:
  opts_knit:
    root.dir: "../"

---

<!-- The the root.dir option, moved the root directory up one folder. 

Because my file is saved in my /hw subfolder, and the document renders using the directory of where the file is saved, not the project directory. When I run individual code chunks it sets the working directory to the directory where the .Rproj is stored (which is saved in the CE594R_25_Sonntag folder). I want the code to work both when rendering and for running individual chunks. So, I used the root.dir to set the working directory to the CE594R_25_Sonntag folder For most projects I would recommend saving the .qmd file in the same location as the .Rproj file

Please use relative paths rather than absolute paths. For example, I could put 
root.dir: "C:\\Users\\Sonntag\\Github\\CE594R_25_sonntag" but then the code would not work on someone else's computer. 
-->

knitr options are located here: 

<https://yihui.org/knitr/options/>

```{r}
#| include: False

library(tidyverse)
library(readxl)
library(knitr)

```

### Read in ozone data into a data.frame called ozone.data

- Select "House.Number","Visit","Location","ppm","LOD ppm"
- Filter out the rows with House.Number =='BLANK'
- Create a new column called "O3.below.det" that contains True or FALSE,
  - TRUE values are when the "ppm" values are below the detection limit
  - Values that are below the detection limit have a "<" sign
  - Use the function str_detect() (it's the tidyverse version of grepl in base R)
- Create a new  variable "O3.ppb" from the ppm variable, convert it to a numeric variable, and multiply it by 1000
- Create a new variable called O3.estimate that is the "O3.ppb", or if it is below the detection limit, it is the value of "LOD ppm"*1000
  - We are imputing (substituting) the missing values with the limit of detention as the maximum possible value
  - Use the if_else() function


```{r}

```

### Read in the house.info data, and join it to the ozone.data using 'House.Number', into a new data.frame called ozone.data.info

```{r}

```

### Create a summary table called ozone.data.summary
- Return the total number of measurements, and the number of measurements
- the mean, median, min, and max of "O3.ppb"
- measurements grouped by:
  - Location
  - Type of Air Conditioner
    - Note, because there are spaces in `Type of Air Conditioner` refer to the column using the back apostrophe
    - Or rename it with a name with no spaces
- display the ozone.data.summary in a formatted table using kable or tt

```{r}

```

### Make a wide table with O3.estimate in two separate columns for In and Out locations

Calculate a new variable called the Indoor/Outdoor ratio which is the indoor O3 concentration divided by the outdoor O3 concentration for each visit. 

Display the equation of the Indoor/Outdoor (I/O) ratio using LaTeX in an equation block. Include the subset the 3 in O3. 

We used the following equation to calculate the indoor/outdoor ratio. 

$$
Indoor/Outdoor \: O_3 \: ratio = \frac{Indoor \: O_3 \: Concentration}{Outdoor \: O_3 \: Concentration}
$$

```{r}

```

### Summarize the Indoor/Outdoor O3 ratio by the type of air conditioning type (Central and Evaporative), including the number of measurements, the mean, median, min and max. 


```{r}

```
### Embed the number and the mean indoor/outdoor ration for the two types of air conditioners in a sentence or two. Round the mean values to two decimal points. 

The indoor/outdoor ratio for the `r In.Out.summary$total_n[1]` visits in the Central Air homes was `r round(In.Out.summary$mean[1],2)`.

The indoor/outdoor ratio for the `r In.Out.summary$total_n[2]` visits in the Evaporative Cooler homes was `r round(In.Out.summary$mean[2],2)`.

Knit the document to an html document. Make sure to suppress any warning messages in any of the code-chunks

## Write the ozone.wide to ozone.I.O.csv 

```{r}



```

## HW submission

  - Push the .qmd file to your github repository
  - Provide a link to your repository in the assignment submission

