---
title: "InClass_06_tidyverse"
author: "Darrell Sonntag"
date: "2024-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

## The tidyverse packages

<https://www.tidyverse.org/packages/>

Share tidyverse cheat-sheets in class


# We will be using our family data.frame we used in our InClass_03_exercise 

You can re-create it here. It has a column called 'name', and a column called 'age'.

```{r}
name <- c("Darrell", "Amy", "Elisa","Lydia","Emmy")
age <- c(42, 42, 15, 13, 11)
family <- data.frame(name, age)
```

# Make a vector called "age" that stores the ages of those family members in the same order. 

We will start with dplyr

## Piping

In dplyr the symbol %>% is referred to as a pipe. 

> x %>% f(y) becomes f(x,y)

For example, calculate the mean age of your family

```{r}
mean(family$age)
```

Now do it using the piping function

```{r}
family$age %>%
mean()
```

Another example, create a subset of the family, to just look at the kids. All the kids under age 20

```{r}
kids <- filter(family, age<20)
print(kids)
```

Now do this using the piping function

```{r}
kids <- family %>%
        filter(age<20)
print(kids)
```

The cool thing about piping is that we can built upon it. 
Filter your family to those younger than 20, and just select the names


```{r}
kid_names <- family %>%
                filter(age<20) %>%
                select(name) 

print(kid_names)
```

## introduce joining, bind_rows, bind_cols

create another data.frame called family2 (extended family?)

Have a column called name, and a column called age


```{r}
family2 <- data.frame(name = c('Paul','Carol','Alison','David','Stephanie'),age = c(79,76,46,44,39))
```

Then create a new table by stacking the family and family2 data.frames together using the function bind_rows(). Call it family_extended

```{r}
family_extended <- bind_rows(family,family2)
```

Do the same thing, but use the piping operator %>%

```{r}
family_extended <- family %>%
                   bind_rows(family2)
```

Create a new data.frame called 'city' with each family name, and their 
city. Put the extended family names first

```{r}
city <- data.frame(name = c('Paul','Carol','Alison','David','Stephanie','Darrell','Amy','Elisa','Lydia','Emmy'),city = c(rep('Fruit Heights',2),'Boise',rep('Farmington',2),rep('Mapleton',5)))

```

Create a new data frame, called 'family_extended_city' by 
binding the city to your family_extended using bind_cols()

```{r}
family_extended_city <- family_extended %>%
                        bind_cols(city)

```

Did it work?

Now Create a new data frame, called 'family_extended_city' by 
joining the city to your family_extended using full_join()

```{r}
family_extended_city <- family_extended %>%
                        full_join(city,by='name')

```

The rows don't need to be the same length for a join
Create a data.frame called state.city with a column for city and state. Add at least one extra city state combination.  

```{r}
state.city <- data.frame(city=c('Mapleton','Fruit Heights','Farmington','Provo','Boise'), state = c(rep('UT',4),'ID'))
```

Join family_extended_city with state.city


```{r}
family_extended_city_state <- family_extended_city %>%
                        left_join(state.city,by='city')

```
Which join to use? (left, right, inner, or full?)

left

Which column should did join by?

city


Calculate a new column called 'year_born' from the age

```{r}

family_extended_city_state['year_born'] <- 2023-family_extended_city_state['age']

```

Now calculate 'year_born'  the tidyverse way using mutate

```{r}
family_extended_city_state <- mutate(family_extended_city_state, year_born = 2023-age)
```

### if else statements

We could have used an if_else() statement to add the State name.
Use that approach

```{r}
family_extended_city_state <- family_extended_city %>%
                              mutate(state=if_else(city=='Boise','ID','UT'))

```

I often like to use joins, rather than using if_else statements, especially if we had lots of city and state combinations. I think a table is easier to use, then a complex if statement. 

##

But there are often cases, where if_else statements make more sense. 

Using an if_else statement to add a column called life_stage with either adult, adolescent, or child as values. Have the life_stage = 'child' if their age is <13, teenager < 20, and adult for 20 and older.

```{r}
family_extended_city_state <- family_extended_city_state %>%
                              mutate(life_stage = if_else(age<13,'child',
                                                          if_else(age<20,'teenager','adult')))
                                                                  
```


Now go from family, bind family2, join  city and the the state, and add columns year_born, life_stage all in one step using piping %>%

```{r}
family_extended_city_state <- family %>%
                              bind_rows(family2) %>%
                              full_join(city,by='name') %>%
                              left_join(state.city,by='city') %>%
                              mutate(year_born = 2023-age) %>%
                              mutate(life_stage = if_else(age<13,'child',
                                                     if_else(age<20,'teenager','adult')))
                              
                          
```

Other tips, you can arrange rows and relocate columns, 
Arrange the data by year_born and then age
Order the columns, using the select statement by name,age,year_born,life_stage,city,state


```{r}
family_extended_city_state <- family %>%
                              bind_rows(family2) %>%
                              full_join(city,by='name') %>%
                              left_join(state.city,by='city') %>%
                              mutate(year_born = 2023-age) %>%
                              mutate(life_stage = if_else(age<13,'child',
                                                     if_else(age<20,'teenager','adult'))) %>%
                              arrange(year_born,age) %>%
                              select(name,age,year_born,life_stage,city,state)

```

## What's cool about tidyverse?

- Vectorized functions (don't need to use for or while loops) 
  - [What are vectorized functions](https://smogdr.github.io/edar_coursebook/rprog4.html)
- Functions are human readable (join, arrange, mutate)
- Can write compact code that you can follow
- Don't need to create intermediate files
- Don't need to have such long names for my files 
    - For example, family_extended_city_state isn't needed because there are no longer a bunch of intermediate files

## Summarize functions

Use the summary function to summarize the family_textended_city_state

```{r}
summary(family_extended_city_state)
```


Calculate the mean using the summarize function

```{r}

mean_age <- summarize(family_extended_city_state,
                      mean_age = mean(age))

print(mean_age)
```
Or we could have just calculated the mean by using the mean

```{r}

mean_age <- mean(family_extended_city_state$age)

print(mean_age)
```
Q: What's the difference?
A: 1) summarize returns a data.frame, the mean function returns a value
   2) summarize can calculate summaries using multiple functions and multiple columns


Now calculate the mean age, mean year born, and the number of observations using the summarize statement
```{r}

mean_age <- summarize(family_extended_city_state,
                      mean_age = mean(age), mean_year_born = mean(year_born), people = n())

print(mean_age)
```

Calculate mean, min, and max age

```{r}
family_sum <- family_extended_city_state %>%
                summarize(
                     mean_age = mean(age),min_age=min(age),max_age=max(age))
print(family_sum)
```
## Grouping calculate the average, min and max, and count n(), by by life_stage

Use group_by()

```{r}
family_sum <- family_extended_city_state %>%
                group_by(life_stage) %>%
                summarize(
                     mean_age = mean(age),min_age=min(age),max_age=max(age), count=n())
print(family_sum)
```

You can group by multiple variables,

group by life stage and city
```{r}
family_sum <- family_extended_city_state %>%
                group_by(life_stage,city) %>%
                summarize(
                     mean_age = mean(age),min_age=min(age),max_age=max(age), count=n())
print(family_sum)
```

Note about working directories in Rmarkdown

The working directory within each code chunk is by default the location where the .RMD file is saved.

Info on the working directory in .rmd is here:

<https://bookdown.org/yihui/rmarkdown-cookbook/working-directory.html>

## read in air_passengers.csv data saved in the data folder of the public repository

```{r}
air_passengers <- read_csv("..\\..\\CE594R_data_science_R\\data\\air_passengers.csv")
```
What if you wanted to calculate the mean the passengers by month?

```{r}

air_passengers_summary <- air_passengers %>%
                          summarize(Jan_mean = mean(Jan),Feb_mean= mean(Feb))
```

What if you wanted to calculate other summary stats, such as the min, max, sd for each month?


Let's first change the format of the table to a long format using pivot_longer() 
See the tidyr cheat sheet!

```{r}

?pivot_longer()

air_passengers_long <- pivot_longer(air_passengers,cols=2:13,names_to='month',values_to='passengers')
```

Now calculate the max,min, max, and sd for each month

```{r}

air_passengers_summary <- air_passengers_long %>%
                          group_by(month) %>%
                          summarize(mean = mean(passengers), min=min(passengers),max=max(passengers),sd=sd(passengers))
```

It worked!... except I don't like how the months aren't ordered right.

How can we fix that?
Mutate month using factor(), specify the levels and order of the months

```{r}

air_passengers_summary <- air_passengers_long %>%
                          mutate(month = factor(month,levels=c('Jan', 'Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec')),ordered=T) %>%
                          group_by(month) %>%
                          summarize(mean = mean(passengers), min=min(passengers),max=max(passengers),sd=sd(passengers))
```


